The Small World Network Effect in Software Project Teams \author{Kevin Peterson\\ \small \texttt{pete1968@umn.edu} }

### 2013-12-06 paper/main.tex
One more proofreading pass.

\subsection{Hypothesis}
The formalization of the ``small world'' quality of a graph can be expressed by the \textit{Small World Quotient}, or $Q$\cite{watts1999small,watts1998collective}. The value $Q$ is calculated by dividing the clustering factor of graph by the average path length between nodes. Uzzi explored this value in graphs of Broadway productions\cite{uzzi2005collaboration}. His findings suggest that the success of Broadway productions is greatest when teams have a good mix of new and familiar members. Productions are less likely to be successful when team members are very familiar with each other. Also, probability of success drops much the same way when there is little familiarity between team members. This indicates that there is an optimal ratio of new and familiar members on a team -- or, an optimal value of $Q$ when related to performance.

@@

\subsection{Collection and Storage}
Data was collected using the freely available FLOSSmole\cite{floss2006} datasets of free and open source projects. Two datasets were analyzed -- data from from the popular open source project site \textit{Freecode}\footnote{http://freecode.com/}, and another from open source software repository \textit{SourceForge}\footnote{http://sourceforge.net/}. The Freecode dataset contained data up to September, 2013, while the SourceForge dataset was slightly older with data gathered on June 2009. Both datasets were chosen because they met two basic criteria: (1) they contained an exhaustive list of projects and contributors to those projects, and (2) they provided some sort of quantitative measure of project ``popularity." This popularity metric will be further described in sections below. Results were analyzed using the Python graph processing packaged NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

@@

\subsubsection{Graph Structure}
For each dataset, we first considered the dataset to be one large graph. It was observed that the contributor graph was ``disconnected,'' meaning that there was not a path between all possible nodes. From this large disconnected graph, we extracted a set of connected subgraphs, where each connected subgraph was analyzed. Metrics were then computed for each of these subgraphs. In order to correctly compute the metrics below, a minimum subgraph size was required. Specifically, many of the graph metrics of interest measured ``triangles'' of nodes and their connections. This implies a subgraph of {$>= 3$} nodes is needed. Subgraphs of {$< 3$} nodes were not considered for analysis.

@@

A bipartite graph projection is necessary to further analyze the graph. A bipartite projection involves taking the disjoint node sets $P$ and $C$, and representing the graph as relationships between only one of those sets of nodes. A projection onto the contributor nodes, or $C$, of figure \ref{fig:example_bipartite_graph} is show in figure \ref{fig:example_bipartite_projection_graph}. Here, contributors are connected directly, with edges occurring if a two contributors share one or more common projects. It is important to note that the projection could have been done in terms of the projects, instead of the contributors. This would have led to a graph with nodes of projects $P$, each being linked by sharing a common contributor. Projection is a common way of representing these bipartite graphs\cite{newman2001scientific}, although it is known that bipartite projections are lossy compared to the data represented in the original graph\cite{zhou2007bipartite}.\\

@@

\noindent\\\textit{\textbf{Clustering Coefficient - Average ($\overline{C^{\lambda}}$)}}\\
Another approach to clustering is to take the average of each node's local clustering coefficient\cite{watts1998collective}.

@@

From this \textit{Local Clustering Coefficient} $C^{\lambda}_i$, we can then take the average over the entire set of nodes in the graph.
\[ \overline{C^\lambda} = \frac{1}{|P|}\sum^{|P|}_{i=1}C^{\lambda}_i \]

@@

\noindent\\\textit{\textbf{Small World Quotient ($Q$)}}\\
The definition of a \textit{small world} graph can be formalized by the following equations\cite{humphries2008network,uzzi2005collaboration}.

@@

Finally, the ratio of the above calculations yields the small world quotient, or $Q$, of the graph
\[Q = \frac{\gamma^{\Delta}_g }{\lambda_g} \]

@@

\noindent\\\textit{\textbf{Performance}}\\
To further explore the data, a measurement of project performance was needed. As the explored datasets consisted entirely of open source projects, research indicates that \textit{quality}, \textit{use}, \textit{user satisfaction}, and \textit{impact} are possible metrics to measure performance or \textit{success}\cite{crowston2003defining}.\\

For both datasets, we base our performance metric on \textit{use}. For Freecode, we use the \textit{Popularity Score} metric, which is calculated by the following formula\footnote{http://help.freecode.com/kb/statistics/how-do-you-measure-a-projects-popularity}:

@@

For the SourceForge dataset, performance is based on the repository's internal \textit{rank} metric. This metric ranks each project from 1 to the total number of projects in SourceForge ($P_{sf}$), or ${1 <= rank <= |P_{sf}|}$. At the time of this writing, no formula for the \textit{rank} metric could be located for citation.  Note that there was no attempt to normalize the two performance metrics between datasets. Because of this, no cross-dataset measurements can be made, as the performance metric is only valid in the context of the enclosing dataset.

\subsubsection{Qualitative Analysis}
So far we have explored, via data analysis, how closely project collaborators are connected to one another, and what correlations we can derive from this analysis. We have said little as to why these connections are made, how they are maintained, and what exactly a connection means at a personal level. Via qualitative analysis, we will attempt to enrich our data analysis by providing deeper context.\\
A connection between two people is much more than an edge between two nodes in a graph. It has been asserted that people tend to make connections with people with whom they share similarities\cite{mcpherson2001birds}. This idea of homophily, however, needs to be further expanded before we can make assertions about correlations to performance. It has been shown that demographic homophily is complex to measure, correlates little to performance outcomes, and in fact correlates negatively\cite{reagans2004make,lawrence1997perspective}. For our purposes, it is much more appropriate to examine how teams best leverage similarities of knowledge and information, as the flow of these assets can be a key driver to success\cite{nissen2002extended}.\\
In the same way we can explore the role of team familiarity. The two types of familiarity that we will explore are \textit{member} familiarity and \textit{task/project} familiarity\cite{harrison2003time}. These metrics, although not captured in data, could be important factors in \textit{how} teams form, and thus help us to understand \textit{why} teams perform the way that they do.\\

@@

A mean value of $Q$ was observed to be $> 1$ for both data sets, which meets the criteria for a small world network\cite{humphries2008network}. The median, however, is 1.0 in both cases, which does not match the small world network criteria. A possible explanation of this is the power law distribution of $Q$ observed in both data sets (figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}). This indicates that a large proportion of $Q$ values are around 1.\\

$C^\Delta$ and $\overline{C^\lambda}$ both indicate values approaching 1. This observation indicates a high degree of clustering\cite{watts1998collective}. Exploring this phenomenon further, we can derive that most graphs resemble the middle or the bottom graph in figure \ref{fig:example_graphs}, with relatively few resembling the top graph.\\

Note that $P_c$ values are relatively small compared to $P$ and $C$ values. It would be intuitive to assume that {$N \times P_c \approx C$} -- in other words, if we split the entire graph into connected subgraphs, the number of subgraphs ($P_c$) multiplied by the average number of nodes ($N$) per subgraph should approximately equal the total number of contributors ($C$). This, however, is not the case. In fact, $P_c$ is notably smaller than expected. Recall that during analysis, subgraphs of {$< 3$} nodes were not considered. We can assume then that a large number of subgraphs are very small, with only one or two nodes. Further analysis is needed to fully understand the implications of this.

To validate the hypothesis, we expect to find that intermediate values of $Q$ will be correlated with more popular projects. To explore this possibility, we first compare the popularity given $Q$ values as shown in figures \ref{fig:q_fc_graph} and \ref{fig:q_sf_graph}. The $Q$ values do display the inverse U-shaped characteristic as compared to popularity, which aligns with the findings of Uzzi \textit{et. al}\cite{uzzi2005collaboration}.

@@

\noindent\\\textit{\textbf{Priority, focus, and timeline are strong drivers.}}\\
Focus was specifically called out as an indicator of performance in one interview. Clear goals and a common objective were positive influencer, while continuously shifting focus was a negative influence. Research in this area suggests that creative teams have ``goals that are clear and compelling, but also open and challenging.''\cite{isaksen2002climate}

Timeline and priority were discussed as strong factors in team success. Team building could be thought of as function of the timeline -- meaning that as the timeline of the project increased, more time could be devoted to team building activities.

@@

\section{Future Research}
Social networks, interpersonal relationships, and software development are all complex topics. We have explored a narrow slice of these in context, and there are some important areas of future exploration. First, the contributor graph contains a large amount of subgraphs, and as a whole appears less clustered than some other observed social network graphs -- consistent with Madey's findings\cite{madey2002open} on SourceForge data. As noted in our analysis, a large number of these subgraphs have only one or two contributors. This implies that a large number of contributors work alone and on a single project. We do not, with the current analysis, know how (or if) this degrades the quality of our findings.\\

A further area of study would be how our findings relate to previously defined \textit{success factors} in projects\cite{cooke2002real}. Also, we do not know directly \textit{how} intermediate values of $Q$ influence performance -- we can only observe a quantitative and qualitative correlation. We make a tacit equivalence to clustering and knowledge sharing, but we have not directly observed that in our analysis. Research does suggest that social factors are knowledge sharing motivators\cite{hendriks1999share}, but how that relates specifically to our work is unexplored here.\\

>>>>

\subsection{Hypothesis}
The formalization of the ``small world'' quality of a graph can be expressed by the \textit{Small World Quotient}, or $Q$\cite{watts1999small,watts1998collective}. The value $Q$ is calculated by dividing the clustering factor of graph by the average path length between nodes. Uzzi explored this value in graphs of Broadway productions\cite{uzzi2005collaboration}. His findings suggest that the success of Broadway productions is greatest when teams have a good mix of new and familiar members. Productions are less likely to be successful when team members are very familiar with each other. Also, probability of success drops much the same way when there is little familiarity between team members. This indicates that there is an optimal ratio of new and familiar members on a team -- or, an optimal value of $Q$ when related to performance. Thus, we formulate our hypothesis:\\

@@

\subsection{Collection and Storage}
To explore our hypothesis, software project data was collected and analyzed. This data was collected using the freely available FLOSSmole\cite{floss2006} datasets of free and open source projects. Two datasets were analyzed -- data from from the popular open source project site \textit{Freecode},\footnote{http://freecode.com/} and another from open source software repository \textit{SourceForge}.\footnote{http://sourceforge.net/} The Freecode dataset contained data up to September 2013, while the SourceForge dataset was slightly older with data gathered up to June 2009. Both datasets were chosen because they met two basic criteria: (1) they contained an exhaustive list of projects and contributors to those projects, and (2) they provided some sort of quantitative measure of project ``popularity." This popularity metric will be further described in sections below. Results were analyzed using the Python graph processing package NetworkX\cite{hagberg-2008-exploring}. All source code pertaining to the collection and analysis can be found at \url{https://github.com/kevinpeterson/small-world-effect-research}.

@@

\subsubsection{Graph Structure}
For each dataset, we first considered the dataset to be one large graph. It was observed that the contributor graph was ``disconnected,'' meaning that there was not a path between all possible nodes. From this large disconnected graph, we extracted a set of connected subgraphs, where each connected subgraph was analyzed. Metrics were then computed for each of these subgraphs. In order to correctly compute the metrics below, a minimum subgraph size was required. Specifically, many of the graph metrics of interest measured ``triangles'' of nodes and their connections. This implies a subgraph of {$>= 3$} nodes is needed. Because of this, subgraphs of {$< 3$} nodes were not considered for analysis.

@@

A bipartite graph projection is necessary to further analyze the graph. A bipartite projection involves taking the disjoint node sets $P$ and $C$, and representing the graph as relationships between only one of those sets of nodes. A projection onto the contributor nodes, or $C$, of figure \ref{fig:example_bipartite_graph} is show in figure \ref{fig:example_bipartite_projection_graph}. Here, contributors are connected directly, with edges occurring if a two contributors share one or more common projects. It is important to note that the projection could have been done in terms of the projects, instead of the contributors. This would have led to a graph with nodes of projects $P$, each being linked by sharing a common contributor. Projection is a common way of representing these bipartite graphs\cite{newman2001scientific}, although it is known that bipartite projections are lossy compared to the data represented in the original graph\cite{zhou2007bipartite}.

@@

\noindent\\\textit{\textbf{Clustering Coefficient - Average ($\overline{C^{\lambda}}$)}}\\
Another approach to clustering is to take the average of each node's local clustering coefficient\cite{watts1998collective}. To calculate the local clustering coefficient of a node, we follow these steps:

@@

From this \textit{Local Clustering Coefficient} $C^{\lambda}_i$, we can then take the average over the entire set of nodes in the graph.
\[ \overline{C^\lambda} = \frac{1}{|P|}\sum_{i \in P}C^{\lambda}_i \]

@@

\noindent\\\textit{\textbf{Small World Quotient ($Q$)}}\\
The definition of a \textit{small world} graph can be formalized by the following equations\cite{humphries2008network,uzzi2005collaboration}:

@@

Finally, the ratio of the above calculations yields the small world quotient, or $Q$, of the graph.
\[Q = \frac{\gamma^{\Delta}_g }{\lambda_g} \]

@@

\noindent\\\textit{\textbf{Performance}}\\
To further explore the data, a measurement of project performance was needed. As the explored datasets consisted entirely of open source projects, research indicates that \textit{quality}, \textit{use}, \textit{user satisfaction}, and \textit{impact} are each possible metrics to measure performance or \textit{success}\cite{crowston2003defining}.\\

For both datasets, we based our performance metric on \textit{use}. For Freecode, we used the \textit{Popularity Score} metric, which is calculated by the following formula:\footnote{http://help.freecode.com/kb/statistics/how-do-you-measure-a-projects-popularity}

@@

For the SourceForge dataset, performance was based on the repository's internal \textit{rank} metric. This metric ranks each project from 1 to the total number of projects in SourceForge ($P_{sf}$), or ${1 <= rank <= |P_{sf}|}$. At the time of this writing, no formula for the \textit{rank} metric could be located for citation.  Note that there was no attempt to normalize the two performance metrics between datasets. Because of this, no cross-dataset measurements can be made, as the performance metric is only valid in the context of the enclosing dataset.

\subsubsection{Qualitative Analysis}
So far we have explored, via data analysis, how closely project collaborators are connected to one another, and what correlations we can derive from this analysis. We have said little as to why these connections are made, how they are maintained, and what exactly a connection means at a personal level. Via qualitative analysis, we will attempt to enrich our data analysis by providing deeper context.

A connection between two people is much more than an edge between two nodes in a graph. It has been asserted that people tend to make connections with people with whom they share similarities\cite{mcpherson2001birds}. This idea of homophily, however, needs to be further expanded before we can make assertions about correlations to performance. It has been shown that demographic homophily is complex to measure, correlates little to performance outcomes, and in fact correlates negatively\cite{reagans2004make,lawrence1997perspective}. For our purposes, it is much more appropriate to examine how teams best leverage similarities of knowledge and information, as the flow of these assets can be a key driver to success\cite{nissen2002extended}.

In the same way we can explore the role of team familiarity. The two types of familiarity that we will explore are \textit{member} familiarity and \textit{task/project} familiarity\cite{harrison2003time}. These metrics, although not captured in data, could be important factors in \textit{how} teams form, and thus help us to understand \textit{why} teams perform the way that they do.

@@

A mean value of $Q$ was observed to be $> 1$ for both data sets, which meets the criteria for a small world network\cite{humphries2008network}. The median, however, is 1.0 in both cases, which does not match the small world network criteria. A possible explanation of this is the power law distribution of $Q$ observed in both data sets (figures \ref{fig:q_fc_distribution} and \ref{fig:q_sf_distribution}). This indicates that a large proportion of $Q$ values are around 1.

$C^\Delta$ and $\overline{C^\lambda}$ both indicate values approaching 1. This observation indicates a high degree of clustering\cite{watts1998collective}. Exploring this phenomenon further, we can derive that most graphs resemble the middle or the bottom graph in figure \ref{fig:example_graphs}, with relatively few resembling the top graph.

Note that $P_c$ values are relatively small compared to $P$ and $C$ values. It would be intuitive to assume that {$N \times P_c \approx C$} -- in other words, if we split the entire graph into connected subgraphs, the number of subgraphs ($P_c$) multiplied by the average number of nodes ($N$) per subgraph should approximately equal the total number of contributors ($C$). This, however, is not the case. In fact, $P_c$ is notably smaller than expected. Recall that during analysis, subgraphs of {$< 3$} nodes were not considered. We can assume then that a large number of subgraphs are very small, with only one or two nodes. Further analysis is needed to fully understand the implications of this observation.

To validate the hypothesis, we expect to find that intermediate values of $Q$ will be correlated with more popular projects. To explore this possibility, we first compare the popularity given $Q$ values as shown in figures \ref{fig:q_fc_graph} and \ref{fig:q_sf_graph}. The $Q$ values do display the inverse U-shaped characteristic as compared to popularity, which aligns with the findings of Uzzi\cite{uzzi2005collaboration}.

@@

\noindent\\\textit{\textbf{Priority, focus, and timeline are strong drivers.}}\\
Focus was specifically called out as an indicator of performance in one interview. Clear goals and a common objective were positive influencers, while continuously shifting focus was a negative influence. Research in this area suggests that creative teams have ``goals that are clear and compelling, but also open and challenging.''\cite{isaksen2002climate}

Timeline was discussed as a strong factor in team success. Team building could be thought of as function of the timeline -- meaning that as the timeline of the project increased, more time could be devoted to team building activities.

@@

\section{Future Research}
Social networks, interpersonal relationships, and software development are all complex topics. We have explored a narrow slice of these in context, and there are some important areas of future exploration. First, the contributor graph contains a large amount of subgraphs, and as a whole appears less clustered than some other observed social network graphs -- consistent with Madey's findings\cite{madey2002open} on SourceForge data. As noted in our analysis, a large number of these subgraphs have only one or two contributors. This implies that a large number of contributors work alone and on a single project. We do not, with the current analysis, know how (or if) this degrades the quality of our findings.

A further area of study would be how our findings relate to previously defined \textit{success factors} in projects\cite{cooke2002real}. Also, we do not know directly \textit{how} intermediate values of $Q$ influence performance -- we can only observe a quantitative and qualitative correlation. We make a tacit equivalence to clustering and knowledge sharing, but we have not directly observed that in our analysis. Research does suggest that social factors are knowledge sharing motivators\cite{hendriks1999share}, but how that relates specifically to our work is unexplored here.
